Spirals launched by a planet in a shadowed disk

### 2024-04-14 src/tex/ms.tex
Start results

\section{Methods}
\label{sec:methods}
\subsection{Simulation setup}
\label{subsec:setup}
We use the magneto-hydrodynamics code Athena++\footnote{\url{https://github.com/PrincetonUniversity/athena}{https://github.com/PrincetonUniversity/athena}} 
\citep{stone2020} version 21.0 \citep{athena++developmentteam2021} to simulate a 2D gas disk. The simulation was set up in cylindrical coordinates with 149
cells in the $\ln{r}$ direction with $r \in [0.4, 2.5]$ and 512 cells in the $\phi \in [-\pi, \pi]$ direction. The inner and outer boundaries have a user-defined fixed boundary
condition and the azimuthal boundary has a periodic boundary condition to ensure material can orbit the central star indefinitly. The dimensionality of the problem
is set up so that all masses are in $M_\odot$, distances are in AU, and times are in year/$2\pi$. This means that a planet placed in a circular orbit at $r=1$
will have an orbital period of $2\pi$. We chose a planet mass of $10^{-4}$ in order to create a detectable effect in the presence of noise from the shadow and
to not stray too far from the linear spiral regime.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5in]{figures/sound_speed.pdf}
        \caption{Sound speed as a function of azimuthal angle. Here we show how the sound speed is prescribed in our two science cases. The cases
        share a minimum and maximum sound speed, but the transitions between these values are very different.}    
    \end{center}
    \label{fig:cs}
\end{figure}

\begin{figure}
    % \includegraphics[width=0.5\textwidth]{}
    \caption{Initial gas temperature maps for our three setups. {\bf Left:} In the case where there is no shadow the temperature is azimuthally symmetric.
    {\bf Center:} In our ``narrow'' case we see a deep shadow covers a narrow region of the disk. This is analogous to a thin, optically thick inner disk.
    {\bf Right:} In our ``wide'' case we see a much shallower shadow that covers a significant portion of the disk. In this case $a,b=0$ and the third line of
    Equation \ref{eq:H} becomes a sine function for all $\phi$.
    }
    \label{fig:setup}
\end{figure}

@@

>>>>

\section{Methods}
\label{sec:methods}
\subsection{Simulation setup}
\label{subsec:setup}
We use the magneto-hydrodynamics code Athena++\footnote{\url{https://github.com/PrincetonUniversity/athena}} 
\citep{stone2020} version 21.0 \citep{athena++developmentteam2021} to simulate a 2D gas disk. The simulation was set up in cylindrical coordinates with 149
cells in the $\ln{r}$ direction with $r \in [0.4, 2.5]$ and 512 cells in the $\phi \in [-\pi, \pi]$ direction. The inner and outer boundaries have a user-defined fixed boundary
condition and the azimuthal boundary has a periodic boundary condition to ensure material can orbit the central star indefinitly. The dimensionality of the problem
is set up so that all masses are in $M_\odot$, distances are in AU, and times are in year/$2\pi$. This means that a planet placed in a circular orbit at $r=1$
will have an orbital period of $2\pi$. We chose a planet mass of $10^{-4}$ in order to create a detectable effect in the presence of noise from the shadow and
to not stray too far from the linear spiral regime.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5in]{figures/sound_speed.pdf}
        \caption{Sound speed as a function of azimuthal angle. Here we show how the sound speed is prescribed in our two science cases. The cases
        share a minimum and maximum sound speed, but the transitions between these values are very different.}    
    \end{center}
    \script{plot_sound_speed.py}
    \label{fig:cs}
\end{figure}

\begin{figure*}
    \includegraphics[width=1.0\textwidth]{figures/initial_temperature.pdf}
    \caption{Initial gas temperature maps for our three setups. Note that the radial axis is scaled logarithmically. {\bf Left:} In the case where there is no shadow the temperature is azimuthally symmetric.
    {\bf Center:} In our ``narrow'' case we see a near-constant shadow covers a narrow region of the disk. This is analogous to a thin, optically thick inner disk.
    {\bf Right:} In our ``wide'' case we see a much shallower shadow that covers a significant portion of the disk. In this case $a,b=0$ and the third line of
    Equation \ref{eq:H} becomes a sine function for all $\phi$.
    }
    \label{fig:setup}
    \script{plot_initial.py}
\end{figure*}

@@

\subsection{Simulation run}
\label{subsec:sim_run}

We ran each of the five simulations for 100 orbits ($t = 50\pi$) and output the simulation results as an HDF5 file every
quarter orbit ($\Delta t = \frac{\pi}{2}$). The end state of each simulation is shown in Figure \ref{fig:end}.

\begin{figure*}
    \includegraphics[width=1.0\textwidth]{figures/final_density.pdf}
    \caption{Final gas density anomaly maps for our five simulations. Note the patterns are dominated by
    the rings created by the shadow. The radial axis is scaled logarithmically.
    }
    \label{fig:end}
    \script{plot_final.py}
\end{figure*}

\subsection{Isolating planetary effects}
It is clear in Figure \ref{fig:end} that the major deviations in our shadowed disk are due to the shadow rather than the planet. Programically determining
the location of the planet's spiral peak is a non-trivial task. First, we determine the density anomaly $\Delta \rho = \rho - \rho_0$. In the regime where all perturbations
are small we expect that
\begin{equation}
    \Delta \rho_{\rm planet} = \Delta \rho - \Delta \rho_{\rm shadow}
    \label{eq:rho_planet}
\end{equation}

$\Delta \rho_{\rm shadow}$ is determined from our shadow-only simulations (\texttt{narrow\_without} and \texttt{wide\_without}).

Now that we have a density anomaly map we need to determine the location of the spiral peak (in $\phi$) as a function of $r$. Absent of shadow perturbations
(i.e., in the \texttt{no\_shadow} case or if the subtraction in Equation \ref{eq:rho_planet} was perfect) this would be as simple as
finding the value of $\phi$ that maximizes $\rho(\phi,r)$ for any given $r$. However, this is not true in the general case as there are additional
effects that are only found in simulations containing both the planet and shadow. In order to ensure we don't find some other maximum we must constrain
the domain of our search along the $\phi$ axis. One constraint we can utilize is that $\phi_{\rm peak}(r=1) = \phi_{\rm planet}$ as that is where the spiral is
launched from. Since the spiral is a continuous function $\phi(r)$ (that is, its pitch angle is defined, see Section \ref{subsec:pitch}) we can assume that
$\phi(r_i) \approx 2\phi(r_{i-1}) - \phi(r_{i-2})$. Therefore, we can restrict our search for $\phi_{\rm peak}$ based on an initial guess. We restrict searches
to within 0.1 radians of this guess.

\section{Results}
\label{sec:results}

\subsection{Spiral Morphology}

\begin{figure*}
    \script{plot_spirals_narrow.py}
    \includegraphics[width=1.0\textwidth]{figures/spirals_narrow.pdf}

    \caption{
        Spiral patterns observed after 99.25 orbits in the narrow shadow case. Shown is the density anomaly for every case.
        The third panel from the left shows the residuals after subtracting the first panel from the second. Plotted on the two
        rightmost panels are the locations identified as being at the peak of the spiral. Note that the color scale on the leftmost
        two panels has been scaled so that all panels can share the same colorbar.
    }
    \label{fig:sp_narrow}
\end{figure*}

\begin{figure*}
    \script{plot_spirals_wide.py}
    \includegraphics[width=1.0\textwidth]{figures/spirals_wide.pdf}
    \caption{ Same as Figure \ref{fig:sp_narrow} but for the wide shadow case.}
    \label{fig:sp_wide}
\end{figure*}
